D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2023-12-17 ootavacuous.tex
Alan Stern December 16 feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\author{
Paul E.~McKenney\\\email{paulmck@kernel.org} \and
Michael Wong\\\email{fraggamuffin@gmail.com} \and
Maged Michael\\\email{maged.michael@gmail.com} \and
The Indefatigible TBD
}
\date{December 15, 2023 (Pre-Tokyo)}
\maketitle{}

@@

	This paper adds constraints from long-standing laws of physics
	and on hardware systems to demonstrate that OOTA cannot occur
	in correctly constructed C++ programs running on correctly
	constructed C++ implementations.

@@

The current authors deviate slightly from this tradition by instead
defining an OOTA cycle as a cycle in sdep $\cup$ rfe, where rfe is the
set of store-to-load links, but excluding intra-thread rfi links, which
are instead considered to be part of sdep.
Although this does shunt even more complexity onto the term
``semantic dependency'', it also enables us to bring to bear
laws of physics and properties of hardware onto correct real-world
C++ implementations.\footnote{
	Thanks to Alan Stern for providing a litmus test that demonstrated
	that we had unwittingly (but productively!) shifted our definition
	to sdep $\cup$ rfe.}

@@

These examples demonstrate a key point: Although in some cases sdep can
be a strict function of the source code, \emph{in general, sdep must be
defined on a per-execution basis}.\footnote{
	Thanks to Alan Stern for the examples in this section and
	in the preceding one.}

@@

Listing~\ref{lst:Causality Test Case 5}
shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
result is to be forbidden.
The \co{herd7} tool finds this OOTA cycle because C++ does not respect
data dependencies, and flags it as such using \co{S8}.

@@

Omniscient C++ implementations might note that the computation on line~9
can only have the value one when the load from \co{x} returns either
zero or one, and both \co{x} and \co{y} are initialized to zero.
Such an implementation act as if line~10's \co{r2} had instead been
the value~1,
Then there would be no semantic dependency between \co{P0()}'s load
and store, and thus no OOTA cycle in a correct real-world C++
implementation.

@@

We are grateful to David Goldblatt, Jade Alglave, Alan Stern, and Peter
Sewell for their careful review of an early draft of this paper and
to John Wickerson for asking Paul for a rant and taking the proffered
rant seriously.
We also owe David Goldblatt a debt of gratitude for his having asked an
insightful question at the right time, his insights on combinations of
OOTA and UB, and for his ``Deathstation 9000'' demonic CPU.
Alan Stern located many unclear statements and gaps in reasoning.
Martin Uecker contributed valuable insights on backwards-propagating UB.
Richard Grisenthwaite patiently explained the architectural constraints
that prevent hardware OOTA.

@@

% @@@ reviewer to do.
% Luc Maranget: Memory models.
% Mark Rutland: ARM memory ordering.
% Will Deacon: ARM hardware, C++, and OOTA.
% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
% Christoph Hellwig: Why slow on BPF memory model.
% Miguel Ojeda: LKMM and Rust.
% Alice Ryhl: LKMM and Rust.
% Ralf Jung: Rust.
% Catalin Marinas: ARM and formal methods (qspinlock).
% Segher Boessenkool:  GCC.
% Dan Lustig: NVIDIA memory ordering.
% Andrea Parri: Memory models, RISC-V hardware.
% Jonas Oberhauser: Memory models.
% Hernan Ponce de Leon: Memory models.
% Palmer Dabbelt: RISC-V hardware.
% Ali Sezgin: P0422R0.
% Tony Tye: P0422R0.
% Other memory-model maintainers, including Peter Zijlstra.
% Derek Williams: PowerPC hardware.
% Hans Boehm: Skeptic, memory models, OOTA.  Suggestions for others?
% Olivier Giroux: Skeptic, memory models, NVIDIA.
% Luke Geeson: Whatever...  Review.
% Ori Lahav: Prior OOTA.
% Viktor Vafeiadis:  Prior OOTA.
% Derek Dreyer: Prior OOTA.
% Brian Demsky: Prior OOTA.

% @@@ Inter-CPU speculation if inter-CPU squashing of speculation.

% Done as of November 30, 2023:
% David Goldblatt
% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
%	Frederic Weisbecker (RCU proteges)
% Michael Wong
% Maged Michael
% John Wickerson: Thank you.  OK for Peter Sewell, Mark Batty, etc.
% Akira Yokosawa: Professional courtesy, perfbook memory models.
% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
% Alan Stern: Memory models, mathematical logic, and variety of hardware.
% Martin Uecker (offered)
% Peter Sewell
% Mark Batty
% Richard Gristenthwaite: ARM hardware.  Posed HW speculation question.

>>>>

\author{
Paul E.~McKenney\\\email{paulmck@kernel.org} \and
Michael Wong\\\email{fraggamuffin@gmail.com} \and
Maged Michael\\\email{maged.michael@gmail.com} \and
Alan Stern\\\email{stern@rowland.harvard.edu} \and
The Indefatigible TBD
}
\date{December 15, 2023 (Pre-Tokyo)}
\maketitle{}

@@

	This paper refactors the commonly used condition for detecting
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and on hardware systems.
	These constraints prevent OOTA from occurring in correctly
	constructed C++ programs running on correctly constructed C++
	implementations in our current universe.

@@

The current authors deviate slightly from this tradition by instead
defining an OOTA cycle as a cycle in sdep $\cup$ rfe (or
\co{acyclic(sdep ; rfe)} in \co{herd7} notation), where rfe is the
set of store-to-load links, but excluding intra-thread rfi links, which
are instead considered to be part of sdep.
Although this does shunt even more complexity onto the term
``semantic dependency'', it also enables us to bring to bear
laws of physics and properties of hardware onto correct real-world
C++ implementations.% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}

@@

These examples demonstrate a key point: Although in some cases sdep can
be a strict function of the source code, \emph{in general, sdep must be
defined on a per-execution basis}.% \footnote{
% 	Thanks to Alan Stern for the examples in this section and
% 	in the preceding one.}

@@

Listing~\ref{lst:Causality Test Case 5}
shows causality test case 5, for which the \co{r1 == r2 == 1 && r3 == 0}
result is to be forbidden.
The \co{herd7} tool finds this OOTA cycle because C++ does not respect
data dependencies, and flags it as such using \co{S8}.
Interestingly enough, there is no state with \co{r3} also shown as
\co{S8}, perhaps indicating that \co{herd7} does not propagate OOTA
values.

@@

Omniscient C++ implementations might note that the computation on line~9
can only have the value one when the load from \co{x} returns either
zero or one, and both \co{x} and \co{y} are initialized to zero.
Such an implementation act as if line~10's \co{r2} had instead been
the value~1, so that the value returned by \co{P0()}'s load could not
affect the value stored.
Then there would be no semantic dependency between \co{P0()}'s load
and store, and thus no OOTA cycle in a correct real-world C++
implementation.

@@

We are grateful to David Goldblatt, Jade Alglave, and Peter
Sewell for their careful review of an early draft of this paper and
to John Wickerson for asking Paul for a rant and taking the proffered
rant seriously.
We also owe David Goldblatt a debt of gratitude for his having asked an
insightful question at the right time, his insights on combinations of
OOTA and UB, and for his ``Deathstation 9000'' demonic CPU.
% Alan Stern located many unclear statements and gaps in reasoning.
Martin Uecker contributed valuable insights on backwards-propagating UB.
Richard Grisenthwaite patiently explained the architectural constraints
that prevent hardware OOTA.

@@

% @@@ reviewer to do.
% Luc Maranget: Memory models.
% Greg Kroah-Hartman and Linus Torvalds: FYI, trouble being caused.
% Christoph Hellwig: Why slow on BPF memory model.
% Miguel Ojeda: LKMM and Rust.
% Alice Ryhl: LKMM and Rust.
% Ralf Jung: Rust.
% Catalin Marinas: ARM and formal methods (qspinlock).
% Segher Boessenkool:  GCC.
% Dan Lustig: NVIDIA memory ordering.
% Andrea Parri: Memory models, RISC-V hardware.
% Jonas Oberhauser: Memory models.
% Hernan Ponce de Leon: Memory models.
% Palmer Dabbelt: RISC-V hardware.
% Ali Sezgin: P0422R0.
% Tony Tye: P0422R0.
% Other memory-model maintainers.
% Derek Williams: PowerPC hardware.
% Hans Boehm: Skeptic, memory models, OOTA.  Suggestions for others?
% Olivier Giroux: Skeptic, memory models, NVIDIA.
% Luke Geeson: Whatever...  Review.
% Ori Lahav: Prior OOTA.
% Viktor Vafeiadis:  Prior OOTA.
% Derek Dreyer: Prior OOTA.
% Brian Demsky: Prior OOTA.

% Done as of November 30, 2023:
% David Goldblatt
% Boqun Feng, Uladzislau Rezki, Neeraj Upadhyay, Joel Fernandes, and
%	Frederic Weisbecker (RCU proteges)
% Michael Wong
% Maged Michael
% John Wickerson: Thank you.  OK for Peter Sewell, Mark Batty, etc.
% Akira Yokosawa: Professional courtesy, perfbook memory models.
% Dan Kelley, Alexei Starovoitov, Mykola Lysenko: FYI.  OOTA dropped
% Alan Jeffrey: Fixed-point insight P0422R0.  Reached out via LinkedIn.
% Jade Alglave: C++ herd model?  Memory models and variety of hardware.
% Alan Stern: Memory models, mathematical logic, and variety of hardware.
% Martin Uecker (offered)
% Peter Sewell
% Mark Batty
% Richard Gristenthwaite: ARM hardware.  Posed HW speculation question.
% Peter Zijlstra.
% Mark Rutland: ARM memory ordering.
% Will Deacon: ARM hardware, C++, and OOTA.
